An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2017-03-29 old_paper_versions/NIPS_Reviews_Rebuttals/rebuttal_reviewers.txt

We thank the reviewers for their thoughtful commentary on our paper.
We can only agree with the comments and appreciate the reviewers
patience in reviewing the paper. It was an early draft. We noticed
most of these issues ourselves in the time since we submitted the paper
and have been working to address them. We discuss those corrections
in detail below.

General:

First we established worst-case lower bounds for references [3] and [4]
of Omega(N) bounds on the number of samples consumed per iteration,
using simple Gaussian examples.

In addition, Section 4 (the theoretical results) has been
substantially revised.  Rather than general appeal to CLT, we include
bounds based on t-statistics (using both the empirical mean and
variance) asymptotics from [A] (below) with explicit constants and an
O(n^-0.5) error bound. We revised the analysis of the test error rate
which now requires no additional assumptions. We include error bounds
on the equilibrium distribution given bounds on transition kernel
error which appeared in [3] and [4], citing those works, under the
appropriate assumptions.

To Reviewers 2, 3, and 4 regarding the Barker function:

You are correct, the logistic (Barker) function is less efficient than
the Metropolis acceptance test with \theta' close to \theta when run
on the full dataset. However, there is no comparable result when these
tests are used on minibatches since they are applied in completely
different ways. Empirically, we find that we can scale proposal step
size or temperature to achieve as small minibatch size as desired
using our corrected Barker test, but this is not true of [3] or [4]
which via the analysis above (and in experiments) still require
Omega(n) samples for some single tests. Thus the approx.  Barker test
is arbitrarily more efficient than tail tests in terms of average data
consumed per generated distinct sample. We have revised the
experiments so show that such gaps are up to 3 orders of magnitude for
logistic regression and the gaussian mixture model.

To Reviewers 1, 2, and 3 regarding the deconvolution (in)stability:

We now describe in detail the construction of the distribution of
X_corr. We explored linear programming and least squares solutions. We
find that the latter allow us to compute finer discrete approximations
to the distribution of X_corr (4000 points). It also gives smaller
absolute (i.e. L1) error. Even though least squares minimizes the L2
norm of the error, the L1 error is also bounded down to 1e-5 by L2
norm minimization. The solution accuracy benefits from fairly strong
L2 regularlization, which also makes the inversion
(i.e. deconvolution) more stable. To ensure positivity of all three
distributions, we impose a stricter bound on the variance of X_norm.
This effectively shifts variance to $X_corr$ by convolving it with
another normal distribution, and guarantees that it is positive.
We computed X_corr with variance bounds of 0.8 and 0.9 both of
which give CDF errors below 1e-4.

To Reviewer 3:

(TODO respond to experiment criticism, but I really can't defend the work as it is ...)

Both Reviewers 1 and 3 correctly identified our implicit assumptions
about Lemma 2. Reviewer_1 is right that we approximated it for large
N, so sampling without replacement approximates sampling with
replacement.  In the case of very large datasets however, it is also
typically more reasonable to treat the dataset as itself a sample
drawn from a still larger, or possibly infinite, dataset (e.g. this
holds for all the examples in our paper), and the instance
"resolution" implies that repeats have probability ~ zero. So
distinctions between sampling with replacement and without are perhaps
moot for these datasets.

To Reviewers 1 and 2:

Thank you for mentioning [BDH], "MCMC Methods for Tall Data." We added
this reference as part of our ongoing revision of the paper and will
substantially address issues raised in [BDH]. In addition, we are
addressing Reviewer 2's concern regarding the difference between ours
and [BDH] Section 6.3. The idea of using a Barker test with minbatches
was proposed in [BDH]. However, a crucial addition in our approach is
the use of an additive correction variable. Thus in our approach,
there is no lower bound on error in our approximate test, and we show
it decreases with O(n^-0.5). In [BDH] there is no correction between
normal and logistic CDFs, and the method has a minimum error set by
the max absolute difference between their CDFs. Another difference is
that the idea proposed in [BDH] was not explored experimentally. In
particular it was not shown that tests can be done with arbitrarily
small minibatches in [BDH].

To Reviewer 7:

We have improved the algorithm to use an adaptive step size rather
than fail on some tests.  Unlike [3] and [4] however, since the amount
of data required for our test scales in proportion to the proposal
step size, the minibatch size has a normal size bound.

[A] Y. Novak, "On Self-Normalized Sums And Student’s Statistic"
Theory of Probability and its Applications, 49(2), 336-344, 2005.

### 2017-03-29 old_paper_versions/ArXiv_October_2016_version/arxiv.tex

\documentclass[twoside]{article} \usepackage{aistats2017}

% If your paper is accepted, change the options for the package
% aistats2017 as follows:
%
%\usepackage[accepted]{aistats2017}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.
\usepackage{amsmath,amsthm,color,graphicx,verbatim,listings,enumitem,float}
\usepackage{amssymb}
\usepackage{natbib}  % for references
\usepackage{amsfonts}
\usepackage{nicefrac}  
\usepackage{booktabs}
\usepackage{microtype}
\usepackage[]{algorithm2e}
\usepackage{caption}
\usepackage{subcaption}
\graphicspath{{figures/}}
\usepackage{setspace}
\lstset{
numbers=left, 
numberstyle=\small, 
numbersep=8pt, 
frame = single, 
language=matlab, 
framexleftmargin=20pt}

\renewcommand{\refname}{}
\newcommand{\simiid}{\overset{\textrm{i.i.d.}}{\sim}}
\DeclareMathOperator*{\argmin}{arg\,min}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}

% This portion of style definition will remove the index number before the 
% without inducing an indent
\makeatletter
\renewcommand\@biblabel[1]{}
\renewenvironment{thebibliography}[1]
     {\section*{\refname}%
      \@mkboth{\MakeUppercase\refname}{\MakeUppercase\refname}%
      \list{}%
           {\leftmargin0pt
            \@openbib@code
            \usecounter{enumiv}}%
      \sloppy
      \clubpenalty4000
      \@clubpenalty \clubpenalty
      \widowpenalty4000%
      \sfcode`\.\@m}
     {\def\@noitemerr
       {\@latex@warning{Empty `thebibliography' environment}}%
      \endlist}
\makeatother

\begin{document}

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

\twocolumn[

\aistatstitle{An Efficient Minibatch Acceptance Test for Metropolis-Hastings}

\aistatsauthor{ Anonymous Author 1 \And Anonymous Author 2 \And Anonymous Author 3 }

\aistatsaddress{ Unknown Institution 1 \And Unknown Institution 2 \And Unknown Institution 3 } ]

\begin{abstract}
  %The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
  %both left and right-hand margins. Use 10~point type, with a vertical
  %spacing of 11~points. The {\bf Abstract} heading must be centered,
  %bold, and in point size 12. Two line spaces precede the
  %Abstract. The Abstract must be limited to one paragraph.
  We
  present a novel Metropolis-Hastings method for large datasets that
  uses small expected-size minibatches of data. Previous work on
  reducing the cost of Metropolis-Hastings tests yield variable data
  consumed per sample, with only constant factor reductions versus
  using the full dataset for each sample.  Here we present a method
  that can be tuned to provide arbitrarily small batch sizes, by
  adjusting either proposal step size or temperature. Our test uses
  the noise-tolerant Barker acceptance test with a novel additive
  correction variable.  The resulting test has similar cost to a normal
  SGD update. Our experiments demonstrate several order-of-magnitude
  speedups over previous work.
\end{abstract}

@@

Markov chain Monte Carlo (MCMC) sampling is a powerful method for computation on
intractable distributions. We are interested in large dataset applications,
where the goal is to sample a posterior distribution $p(\theta \mid x_1, \ldots,
x_N)$ of parameter $\theta$ for large $N$.  The Metropolis-Hastings method (M-H)
generates sample candidates from a proposal distribution $q$ which is in general
different from the target distribution $p$, and decides whether to accept or
reject based on an acceptance test. The acceptance test is usually a Metropolis
test~\citep{Metropolis1953, hastings70}.

Many state-of-the-art machine learning methods, and deep learning in particular,
are based on minibatch updates (such as SGD) to a model.  Minibatch updates
produce many improvements to the model for each pass over the dataset, and have
high sample efficiency.  In contrast, conventional M-H requires calculations
over the full dataset to produce a new sample.  Recent results
from~\citet{cutting_mh_2014} and~\citet{icml2014c1_bardenet14} perform
approximate (bounded error) acceptance tests using subsets of the full dataset.
The amount of data consumed for each test varies significantly from one
minibatch to the next. By contrast,~\citet{conf/uai/MaclaurinA14,TallData15}
perform exact tests but require a lower bound on the parameter distribution across
its domain.  The amount of data reduction depends on the accuracy of this bound,
and such bounds are only available for relatively simple distributions.

Here we derive a new test which incorporates the variability in minibatch
statistics as {\em a natural part of the test} and requires less data per
iteration than prior work. We use a Barker test function~\citep{Barker65}, which
makes our test naturally error tolerant. The idea of using a noise-tolerant
Barker's test function was suggested but not explored empirically
in~\citet{TallData15} section 6.3. But the asymptotic test statistic CDF and the
Barker function are different, which leads to fixed errors for the approach
in~\citet{TallData15}. Here, we show that the difference between the
distributions can be corrected with an additive random variable. This leads to a
test which is fast, and whose error can be made arbitrarily small.

Our test is applicable when the variance (over data samples) of the log
acceptance probability is small enough (less than 1). It's not clear at first why
this quantity should be bounded, but we will show that it is ``natural'' for
well-specified models running Metropolis-Hastings sampling with optimal
proposals~\citep{OptimalScaling01} on a full dataset. When we reduce the amount
of data for the test, the variance goes up. We have to reduce variance in one
of several ways. Either:

@@

In the Metropolis-Hastings method~\citep{gilks1996markov,brooks2011handbook}, a
difficult-to-compute probability distribution $p(\theta)$ is sampled using a
Markov chain $\theta_1,\ldots,\theta_T$. The sample $\theta_{t+1}$ at time $t+1$
is generated using a candidate $\theta'$ from a (simpler) proposal distribution
$q(\theta'\mid \theta_t)$, filtered by an acceptance test. The acceptance test
is usually a Metropolis test. The Metropolis test has acceptance probability:
\begin{equation}\label{eq:traditional}
    \alpha(\theta_t,\theta') = \frac{p(\theta')q(\theta_t \mid \theta')}{p(\theta_t)q(\theta' \mid \theta_t)} \wedge 1
\end{equation}
where $a \wedge b$ denotes $\min(a,b)$.  With probability
$\alpha(\theta_t,\theta')$, we accept $\theta'$ and set $\theta_{t+1} =
\theta'$, otherwise set $\theta_{t+1}=\theta_t$.  The test is often implemented
with an auxiliary random variable $u \sim \mathcal{U}(0,1)$ with a comparison
$u<\alpha(\theta_t,\theta')$; here, $\mathcal{U}(a,b)$ denotes the uniform
distribution on the interval $[a,b]$.  For simplicity, we drop the subscript $t$
for the current sample $\theta_t$ and denote it as $\theta$.

The acceptance test guarantees detailed balance, which means
$p(\theta)p(\theta'\mid\theta) = p(\theta')p(\theta \mid\theta')$, where
$p(\theta'\mid\theta)$ is the probability of a transition from state $\theta$ to
$\theta'$. Here, $p(\theta'\mid\theta)=q(\theta'\mid\theta)\alpha(\theta,\theta')$.
% Daniel: I commented this out because people should know what it means and we
%re-state another detailed balance equation later!
%\begin{equation}\label{eq:detailed_balance2}
%    p(\theta)q(\theta'\mid\theta)\alpha(\theta,\theta') = p(\theta')q(\theta\mid\theta')\alpha(\theta',\theta)
%\end{equation}
This condition, together with ergodicity, guarantees that the Markov chain has a
unique stationary distribution $\pi(\theta) = p(\theta)$. For Bayesian
inference, the target distribution is $p(\theta \mid x_1, \ldots, x_N)$. The
acceptance probability is now:
\begin{equation}\label{eq:acceptance_probability}
    \alpha(\theta,\theta') = 
    \frac{p_0(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta \mid
    \theta')}{p_0(\theta)\prod_{i=1}^N p(x_i \mid \theta)q(\theta' \mid\theta)}
    \wedge 1
\end{equation}
where $p_0(\theta)$ is the prior. Computing samples this way requires all $N$
data points, but this is very expensive for large datasets. To address this
challenge,~\citet{cutting_mh_2014,icml2014c1_bardenet14} perform approximate
Metropolis-Hasting tests using sequential hypothesis testing. Each iteration,
they start with a small minibatch and test whether $\theta'$ should be accepted
based on approximating $u<\alpha(\theta,\theta')$. If the approximate test
cannot decide with sufficient confidence, the minibatch size is increased and
the test repeats. This process continues until a decision. The bounds depend on
either an asymptotic Central Limit Theorem~\citep{cutting_mh_2014} or a
concentration bound~\citep{icml2014c1_bardenet14}. The latter requires direct
bounds on the log likelihood ratio, which for general distributions requires
knowing $p(x_i \mid \theta)$ and $p(x_i \mid \theta')$ for all $N$ samples. In
addition, while both methods show useful reductions in the number of samples
required, they suffer the drawback of resolving small log likelihood ratio
differences between the minibatch and full batch versions. We discuss a
worst-case scenario in Section~\ref{ssec:gaussian_example}.

Following~\citet{icml2014c1_bardenet14}, we write the test
$u<\alpha(\theta,\theta')$ equivalently as $\Lambda(\theta,\theta') >
\psi(u,\theta,\theta')$, where\footnote{Our definitions differ from those
in~\citet{icml2014c1_bardenet14} by a factor of $N$ to simplify our analysis
later.}
\begin{equation}
\begin{split} 
\Lambda(\theta,\theta') & = \sum_{i=1}^N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right) \\  
{\rm and} \quad \psi(u,\theta,\theta') &= \log\left(u\frac{q(\theta'|\theta)p_0(\theta)}{q(\theta|\theta')p_0(\theta')}\right).
\end{split}
\end{equation}
To reduce computational effort, an unbiased estimate of $\Lambda(\theta,\theta')$
based on a minibatch can be used:
\begin{equation}
\Lambda^*(\theta,\theta') = \frac{N}{b}\sum_{i=1}^b \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\end{equation}
Finally, it will be convenient for our analysis to define
$\Lambda_i(\theta,\theta') = N\log(\frac{p(x_i|\theta')}{p(x_i|\theta)})$.
%% \begin{equation}\label{eq:individual_terms}
%% \Lambda_i(\theta,\theta') = N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
%% \end{equation}
Thus, $\Lambda(\theta,\theta')$ is the mean of $\Lambda_i(\theta,\theta')$ over
the entire dataset, and $\Lambda^*(\theta,\theta')$ is the mean of
$\Lambda_i(\theta,\theta')$ over its minibatch.

@@

The posterior is a Gaussian centered on the sample mean $\mu$, and with variance
$1/N$ (i.e., $\mathcal{N}(\mu, 1/N)$). In one dimension, an efficient proposal
distribution has the same variance as the target
distribution~\citep{OptimalScaling01}, so we use a proposal based on
$\mathcal{N}(\theta,1/N)$. It is symmetric
$q(\theta'\mid\theta)=q(\theta\mid\theta')$, and since we assumed a uniform
prior, $\psi(u,\theta,\theta')=\log u$. Our worst-case scenario is specified in
Lemma~\ref{lem:worst_case}.

\begin{lemma}\label{lem:worst_case}
    For the model in Section~\ref{ssec:gaussian_example}, there exists a fixed
    (independent of $N$) constant $c$ such that with probability $\geq c$ over
    the joint distribution of $(\theta, \theta', u)$, the tests
    from~\cite{cutting_mh_2014,icml2014c1_bardenet14} consume all $N$ samples. 
\end{lemma}
\vspace{-1em}
\begin{proof}
See Supplementary Material, Section~\ref{app:worst_case_proof}.
\end{proof}
Similar results can be shown for other distributions and proposals by
identifying regions in product space $(\theta,\theta',u)$ such that the
hypothesis test needs to separate nearly-equal values.  It follows that the
accelerated tests from prior work require at least a constant fraction $\geq c$
in the amount of data consumed per test compared to full-data tests, so their
speed-up is $\le 1/c$. The issue is the use of tail bounds to separate $\Lambda-\psi$
from zero; for certain input/random $u$ combinations, this difference can be
arbitrarily close to zero. We avoid this by using the {\em approximately normal}
variation in $\Lambda^*$ to {\em replace} the variation due to $u$.

\subsection{MCMC Posterior Inference}
There is a separate line of MCMC work drawing principles from statistical
physics. One can apply
Hamiltonian Monte Carlo (HMC)~\citep{mcmc_hamiltonian_2010} methods which
generate high acceptance \emph{and} distant proposals when run on full batches
of data. Recently Langevin Dynamics~\citep{langevin_2011,conf/icml/AhnBW12} has
been applied to Bayesian estimation on minibatches of data. This simplified
dynamics uses local proposals and avoids M-H tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on proposal
step size is severe, and the state space exploration reduces to a random walk.
Full minibatch HMC for minibatches was described in~\citet{sghmc_2014} which
allows momentum-augmented proposals with larger step sizes. However, step sizes
are still limited by the need to run accurately without M-H tests.  By providing
an M-H test with similar cost to standard gradient steps, our work opens the
door to applying those methods with much more aggressive step sizes without loss
of accuracy.

@@

For our new M-H test, we denote the exact and approximate log likelihood ratios
as $\Delta$ and $\Delta^*$, respectively. First, $\Delta$ is defined as
\begin{equation}\label{eq:delta1}
    \Delta(\theta,\theta')  =
    \log \frac{p_0(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta \mid
    \theta')}{p_0(\theta)\prod_{i=1}^N p(x_i \mid \theta)q(\theta' \mid\theta)},
\end{equation}
where $p_0, p$, and $q$ match the corresponding functions within
Equation~(\ref{eq:acceptance_probability}). We separate out terms dependent and
independent of the data $x$ as:
\begin{equation}\label{eq:delta2}
\begin{split}
    \Delta(\theta,\theta') &= \sum_{i=1}^N\log\frac{p(x_i\mid\theta')}{p(x_i\mid\theta)} - \psi(1,\theta,\theta') \\
    & = \Lambda(\theta,\theta') - \psi(1,\theta,\theta').
\end{split}
\end{equation}
A minibatch estimator of $\Delta$, denoted as $\Delta^*$, is
\begin{equation}\label{eq:delta3}
\begin{split}
    \Delta^*(\theta,\theta') &=
\frac{N}{b}\sum_{i=1}^b\log\frac{p(x_i\mid\theta')}{p(x_i\mid\theta)} - \psi(1,\theta,\theta')\\
&=\Lambda^*(\theta,\theta') - \psi(1,\theta,\theta')
\end{split}
\end{equation}
Note that $\Delta$ and $\Delta^*$ are evaluated on the full dataset and a
minibatch of size $b$ respectively. The term $N/b$ means
$\Delta^*(\theta,\theta')$ is an unbiased estimator of $\Delta(\theta,\theta')$.

@@

This result is used in~\citet{Barker65} to define the Barker acceptance test.
%As
%a sanity check, choosing $g(s) = \exp(s) \wedge 1$ --- a function satisfying the
%requirement of Lemma~\ref{lem:detailed_balance} --- produces the classical
%Metropolis acceptance test $\alpha(\theta,\theta') = g(\Delta(\theta,\theta')) =
%\frac{p(\theta')q(\theta \mid \theta')}{p(\theta)q(\theta' \mid \theta)}\wedge
%1$.
%In fact, $g(s) =\exp(s) \wedge 1$ is the optimal acceptance function in
%terms of acceptance rate, since it accepts with probability 1 for $\Delta > 0$.

>>>>

\documentclass{article}

% if you need to pass options to natbib, use, e.g.: \PassOptionsToPackage{numbers, compress}{natbib}
% before loading nips_2016
% to avoid loading the natbib package, add option nonatbib: \usepackage[nonatbib]{nips_2016}

%\usepackage{nips_2016}
% to compile a camera-ready version, add the [final] option, e.g.:
\usepackage[final]{nips_2016}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amssymb}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}

\usepackage{amsmath,amsthm,color,graphicx,verbatim,listings,enumitem}
\usepackage[]{algorithm2e}
\graphicspath{{figures/}}

\lstset{
numbers=left, 
numberstyle=\small, 
numbersep=8pt, 
frame = single, 
language=matlab, 
framexleftmargin=20pt}

\newcommand{\simiid}{\overset{\textrm{i.i.d.}}{\sim}}
\DeclareMathOperator*{\argmin}{arg\,min}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}

\title{An Efficient Minibatch Acceptance Test for Metropolis-Hastings}

\author{
  Haoyu Chen\textsuperscript{*}, Daniel Seita\textsuperscript{*},
  Xinlei Pan\textsuperscript{*}, John Canny\textsuperscript{*}$^{\dag}$ \\
  * University of California, Berkeley, CA \\
  $\dag$ Google Research, Mountain View, CA\\
  \texttt{\{haoyuchen,seita,xinleipan,canny\}@berkeley.edu}
 }

\begin{document}
\maketitle

\begin{abstract}
  We present a novel Metropolis-Hastings method for large datasets that
  uses small expected-size minibatches of data. Previous work on
  reducing the cost of Metropolis-Hastings tests yield variable data
  consumed per sample, with only constant factor reductions versus
  using the full dataset for each sample.  Here we present a method
  that can be tuned to provide arbitrarily small batch sizes, by
  adjusting either proposal step size or temperature. Our test uses
  the noise-tolerant Barker acceptance test with a novel additive
  correction variable.  The resulting test has similar cost to a normal
  SGD update. Our experiments demonstrate several order-of-magnitude
  speedups over previous work.
\end{abstract}

@@

Markov chain Monte Carlo (MCMC) sampling is a powerful method for computation on
intractable distributions. We are interested in large dataset applications,
where the goal is to sample a posterior distribution $p(\theta \mid x_1, \ldots,
x_N)$ of parameter $\theta$ for large $N$.  The Metropolis-Hastings method (M-H)
generates sample candidates from a proposal distribution $q$ which is in general
different from the target distribution $p$, and decides whether to accept or
reject based on an acceptance test. The acceptance test is usually a Metropolis
test~\cite{Metropolis1953, hastings70}.

Many state-of-the-art machine learning methods, and deep learning in particular,
are based on minibatch updates (such as SGD) to a model.  Minibatch updates
produce many improvements to the model for each pass over the dataset, and have
high sample efficiency.  In contrast, conventional M-H requires calculations
over the full dataset to produce a new sample.  Recent results
from~\cite{cutting_mh_2014} and~\cite{icml2014c1_bardenet14} perform
approximate (bounded error) acceptance tests using subsets of the full dataset.
The amount of data consumed for each test varies significantly from one
minibatch to the next. By contrast,~\cite{conf/uai/MaclaurinA14,TallData15}
perform exact tests but require a lower bound on the parameter distribution across
its domain.  The amount of data reduction depends on the accuracy of this bound,
and such bounds are only available for relatively simple distributions.

Here we derive a new test which incorporates the variability in minibatch
statistics as {\em a natural part of the test} and requires less data per
iteration than prior work. We use a Barker test function~\cite{Barker65}, which
makes our test naturally error tolerant. The idea of using a noise-tolerant
Barker's test function was suggested but not explored empirically
in~\cite{TallData15} section 6.3. But the asymptotic test statistic CDF and the
Barker function are different, which leads to fixed errors for the approach
in~\cite{TallData15}. Here, we show that the difference between the
distributions can be corrected with an additive random variable. This leads to a
test which is fast, and whose error can be made arbitrarily small.

Our test is applicable when the variance (over data samples) of the log
acceptance probability is small enough (less than 1). It's not clear at first why
this quantity should be bounded, but we will show that it is ``natural'' for
well-specified models running Metropolis-Hastings sampling with optimal
proposals~\cite{OptimalScaling01} on a full dataset. When we reduce the amount
of data for the test, the variance goes up. We have to reduce variance in one
of several ways. Either:

@@

In the Metropolis-Hastings method~\cite{gilks1996markov,brooks2011handbook}, a
difficult-to-compute probability distribution $p(\theta)$ is sampled using a
Markov chain $\theta_1,\ldots,\theta_T$. The sample $\theta_{t+1}$ at time $t+1$
is generated using a candidate $\theta'$ from a (simpler) proposal distribution
$q(\theta'\mid \theta_t)$, filtered by an acceptance test. The acceptance test
is usually a Metropolis test. The Metropolis test has acceptance probability:
\begin{equation}\label{eq:traditional}
    \alpha(\theta_t,\theta') = \frac{p(\theta')q(\theta_t \mid \theta')}{p(\theta_t)q(\theta' \mid \theta_t)} \wedge 1
\end{equation}
where $a \wedge b$ denotes $\min(a,b)$.  With probability
$\alpha(\theta_t,\theta')$, we accept $\theta'$ and set $\theta_{t+1} =
\theta'$, otherwise set $\theta_{t+1}=\theta_t$.  The test is often implemented
with an auxiliary random variable $u \sim \mathcal{U}(0,1)$ with a comparison
$u<\alpha(\theta_t,\theta')$; here, $\mathcal{U}(a,b)$ denotes the uniform
distribution on the interval $[a,b]$.  For simplicity, we drop the subscript $t$
for the current sample $\theta_t$ and denote it as $\theta$.

The acceptance test guarantees detailed balance, which means
$p(\theta)p(\theta'\mid\theta) = p(\theta')p(\theta \mid\theta')$, where
$p(\theta'\mid\theta)$ is the probability of a transition from state $\theta$ to
$\theta'$. Here, $p(\theta'\mid\theta)=q(\theta'\mid\theta)\alpha(\theta,\theta')$.
% Daniel: I commented this out because people should know what it means and we
%re-state another detailed balance equation later!
%\begin{equation}\label{eq:detailed_balance2}
%    p(\theta)q(\theta'\mid\theta)\alpha(\theta,\theta') = p(\theta')q(\theta\mid\theta')\alpha(\theta',\theta)
%\end{equation}
This condition, together with ergodicity, guarantees that the Markov chain has a
unique stationary distribution $\pi(\theta) = p(\theta)$. For Bayesian
inference, the target distribution is $p(\theta \mid x_1, \ldots, x_N)$. The
acceptance probability is now:
\begin{equation}\label{eq:acceptance_probability}
    \alpha(\theta,\theta') = 
    \frac{p_0(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta \mid
    \theta')}{p_0(\theta)\prod_{i=1}^N p(x_i \mid \theta)q(\theta' \mid\theta)}
    \wedge 1
\end{equation}
where $p_0(\theta)$ is the prior. Computing samples this way requires all $N$
data points, but this is very expensive for large datasets. To address this
challenge,~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform approximate
Metropolis-Hasting tests using sequential hypothesis testing. Each iteration,
they start with a small minibatch and test whether $\theta'$ should be accepted
based on approximating $u<\alpha(\theta,\theta')$. If the approximate test
cannot decide with sufficient confidence, the minibatch size is increased and
the test repeats. This process continues until a decision. The bounds depend on
either an asymptotic Central Limit Theorem~\cite{cutting_mh_2014} or a
concentration bound~\cite{icml2014c1_bardenet14}. The latter requires direct
bounds on the log likelihood ratio, which for general distributions requires
knowing $p(x_i \mid \theta)$ and $p(x_i \mid \theta')$ for all $N$ samples. In
addition, while both methods show useful reductions in the number of samples
required, they suffer the drawback of resolving small log likelihood ratio
differences between the minibatch and full batch versions. We discuss a
worst-case scenario in Section~\ref{ssec:gaussian_example}.

Following~\cite{icml2014c1_bardenet14}, we write the test
$u<\alpha(\theta,\theta')$ equivalently as $\Lambda(\theta,\theta') >
\psi(u,\theta,\theta')$, where\footnote{Our definitions differ from those
in~\cite{icml2014c1_bardenet14} by a factor of $N$ to simplify our analysis
later.}
\begin{equation}
\Lambda(\theta,\theta') = \sum_{i=1}^N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)
\quad {\rm and} \quad 
\psi(u,\theta,\theta') = \log\left(u\frac{q(\theta'|\theta)p_0(\theta)}{q(\theta|\theta')p_0(\theta')}\right).
\end{equation}
To reduce computational effort, an unbiased estimate of $\Lambda(\theta,\theta')$
based on a minibatch can be used:
\begin{equation}
\Lambda^*(\theta,\theta') = \frac{N}{b}\sum_{i=1}^b \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\end{equation}
Finally, it will be convenient for our analysis to define
$\Lambda_i(\theta,\theta') = N\log(\frac{p(x_i|\theta')}{p(x_i|\theta)})$.
%% \begin{equation}\label{eq:individual_terms}
%% \Lambda_i(\theta,\theta') = N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
%% \end{equation}
Thus, $\Lambda(\theta,\theta')$ is the mean of $\Lambda_i(\theta,\theta')$ over
the entire dataset, and $\Lambda^*(\theta,\theta')$ is the mean of
$\Lambda_i(\theta,\theta')$ over its minibatch.

@@

The posterior is a Gaussian centered on the sample mean $\mu$, and with variance
$1/N$ (i.e., $\mathcal{N}(\mu, 1/N)$). In one dimension, an efficient proposal
distribution has the same variance as the target
distribution~\cite{OptimalScaling01}, so we use a proposal based on
$\mathcal{N}(\theta,1/N)$. It is symmetric
$q(\theta'\mid\theta)=q(\theta\mid\theta')$, and since we assumed a uniform
prior, $\psi(u,\theta,\theta')=\log u$. Our worst-case scenario is specified in
Lemma~\ref{lem:worst_case}.

\begin{lemma}\label{lem:worst_case}
    For the model in Section~\ref{ssec:gaussian_example}, there exists a fixed
    (independent of $N$) constant $c$ such that with probability $\geq c$ over
    the joint distribution of $(\theta, \theta', u)$, the tests
    from~\cite{cutting_mh_2014,icml2014c1_bardenet14} consume all $N$ samples. 
\end{lemma}
\vspace{-1em}
\begin{proof}
See Appendix, Section~\ref{app:worst_case_proof}.
\end{proof}
Similar results can be shown for other distributions and proposals by
identifying regions in product space $(\theta,\theta',u)$ such that the
hypothesis test needs to separate nearly-equal values.  It follows that the
accelerated tests from prior work require at least a constant fraction $\geq c$
in the amount of data consumed per test compared to full-data tests, so their
speed-up is $\le 1/c$. The issue is the use of tail bounds to separate $\Lambda-\psi$
from zero; for certain input/random $u$ combinations, this difference can be
arbitrarily close to zero. We avoid this by using the {\em approximately normal}
variation in $\Lambda^*$ to {\em replace} the variation due to $u$.

\subsection{MCMC Posterior Inference}
There is a separate line of MCMC work drawing principles from statistical
physics. One can apply
Hamiltonian Monte Carlo (HMC)~\cite{mcmc_hamiltonian_2010} methods which
generate high acceptance \emph{and} distant proposals when run on full batches
of data. Recently Langevin Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} has
been applied to Bayesian estimation on minibatches of data. This simplified
dynamics uses local proposals and avoids M-H tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on proposal
step size is severe, and the state space exploration reduces to a random walk.
Full minibatch HMC for minibatches was described in~\cite{sghmc_2014} which
allows momentum-augmented proposals with larger step sizes. However, step sizes
are still limited by the need to run accurately without M-H tests.  By providing
an M-H test with similar cost to standard gradient steps, our work opens the
door to applying those methods with much more aggressive step sizes without loss
of accuracy.

@@

For our new M-H test, we denote the exact and approximate log likelihood ratios
as $\Delta$ and $\Delta^*$, respectively. First, $\Delta$ is defined as
\begin{equation}\label{eq:delta1}
    \Delta(\theta,\theta')  =
    \log \frac{p_0(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta \mid
    \theta')}{p_0(\theta)\prod_{i=1}^N p(x_i \mid \theta)q(\theta' \mid\theta)},
\end{equation}
where $p_0, p$, and $q$ match the corresponding functions within
Equation~(\ref{eq:acceptance_probability}). We separate out terms dependent and
independent of the data $x$ as:
\begin{equation}\label{eq:delta2}
\Delta(\theta,\theta') = \sum_{i=1}^N\log\frac{p(x_i\mid\theta')}{p(x_i\mid\theta)} - \psi(1,\theta,\theta') = \Lambda(\theta,\theta') - \psi(1,\theta,\theta').
\end{equation}
A minibatch estimator of $\Delta$, denoted as $\Delta^*$, is
\begin{equation}\label{eq:delta3}
\Delta^*(\theta,\theta') =
\frac{N}{b}\sum_{i=1}^b\log\frac{p(x_i\mid\theta')}{p(x_i\mid\theta)} -
\psi(1,\theta,\theta')=\Lambda^*(\theta,\theta') - \psi(1,\theta,\theta').
\end{equation}
Note that $\Delta$ and $\Delta^*$ are evaluated on the full dataset and a
minibatch of size $b$ respectively. The term $N/b$ means
$\Delta^*(\theta,\theta')$ is an unbiased estimator of $\Delta(\theta,\theta')$.

@@

This result is used in~\cite{Barker65} to define the Barker acceptance test.
%As
%a sanity check, choosing $g(s) = \exp(s) \wedge 1$ --- a function satisfying the
%requirement of Lemma~\ref{lem:detailed_balance} --- produces the classical
%Metropolis acceptance test $\alpha(\theta,\theta') = g(\Delta(\theta,\theta')) =
%\frac{p(\theta')q(\theta \mid \theta')}{p(\theta)q(\theta' \mid \theta)}\wedge
%1$.
%In fact, $g(s) =\exp(s) \wedge 1$ is the optimal acceptance function in
%terms of acceptance rate, since it accepts with probability 1 for $\Delta > 0$.
